### 2023-03-16 paper.tex
work on methods

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}\\
        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
        \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
        These cells can be discarded with \texttt{scbs filter}.
        To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
        The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
        Identified groups of cells (e.~g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
    }
    \label{fig:workflow}
\end{figure*}

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
The same analysis was repeated for scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
For each data set, we thus obtained four methylation matrices:
either 100~kb tiles or VMRs, quantified either by simple averaging or with the shrunken mean of the residuals.
All four matrices were then subjected to PCA, followed by UMAP with parameters min\_dist $=0.2$, init $=$ spca, and n\_neighbors $=\frac{\sqrt{n}}{1.5}$ where $n$ is the total cell number, to adapt to data sets of different cell numbers.

\subsection{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score

we processed mouse CpG methylation data from \citet{luo2017single} with several method combinations and scored each result.
The score evaluates the ability of each method combination to separate cell types in 15-dimensional PCA space, based on their CpG methylation.
Each score varies between 0 and 1, where higher scores reflect better separation.

Our nearest-neighbor-count score $\text{nnc}_k$, is based on the $\Gamma$-score \citep{Kireeva_2014} and is calculated as follows.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cluster as cell $j$.
We denote this count $a^k_j$.
We consider a cell as ``well-assigned'' if $a^k_j>\theta k$, where we have chosen $\theta=0.9$.
In other words, a cell is ``well-assigned'' if at least 90\% of its $k$ nearest neighbors have been assigned to the same class as the cell under consideration.

The score $\text{nnc}_k$ is then simply the fraction of cells in the data set that are well-assigned in this sense.

% WIP
For all intervals (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.

Regions observed in less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.

For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section.
To visualize data, UMAP was performed on the final 15 principal components.

>>>>

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}\\
        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
        \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
        These cells can be discarded with \texttt{scbs filter}.
        To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
        The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
        Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
    }
    \label{fig:workflow}
\end{figure*}

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
The same analysis was repeated for scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
For each data set, we thus obtained four methylation matrices:
either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
All four matrices were then subjected to PCA.
In the case of methylation fractions, we closely followed methods described in \citet{luo2017single}:
We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca", and n\_neighbors $=\frac{\sqrt{n}}{1.5}$ where $n$ is the total cell number, to adapt to data sets of different size.

\subsection{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$.
The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.
